---
title: "Interactive Visualization in R"
subtitle: " & Applications in Business"
author: "[Amine Elouafi] | [401109492] | [elouafi.amine@stud.hs-fresenius.de]"
date: 05/19/2026
format:
  revealjs:
    self-contained: true
    theme: simple
    slide-number: true
    transition: fade
    code-fold: false
    highlight-style: github
    footer: "Data Analysis for Decision-Making | Fresenius University "
bibliography: references.bib
---

## Agenda

1.  What is Interactive Data Visualization?
2.  Why it matters in Business
3.  The Boston Housing Dataset as an example
4.  Live R Demo with `plotly`
5.  Business Applications
6.  Class Exercise

------------------------------------------------------------------------

## What is Interactive Visualization?

> An interactive chart allows to perform actions **zooming, hovering data through three dimensions** to reveal patterns invisible in 2D charts

::::: columns
::: {.column width="50%"}
**Traditional 2D**

-   Bar charts
-   Line graphs
-   Scatter plots
-   ✅ Simple, familiar
-   ❌ Limited to 2 variables at once
:::

::: {.column width="50%"}
**3D Visualization**

-   3D scatter plots
-   Surface plots
-   3D bar charts
-   ✅ Shows 3 variables together
-   ❌ Can distort if used poorly
:::
:::::

------------------------------------------------------------------------

## When Does Interactive 3D Add Value?

**Use Interactive 3D when:**

-   Three continuous variables must be compared simultaneously.
-   Clusters in high-dimensional data need to be revealed.
-   An interactive HTML format is available (rotation purposes).

**Avoid Interactive 3D when:**

-   The chart will be printed as a static image.
-   Only two variables carry meaningful information.
-   The audience is unfamiliar with 3D interpretation.

::: callout-tip
Always prefer **interactive** 3D (HTML) over static 3D (PDF/print)
:::

------------------------------------------------------------------------

## The Boston Housing Dataset

**Source:** @harrison1978hedonic — 506 census tracts in Boston

**Available in R:** `MASS` package (no download needed!)

```{r}
library(MASS)
data("Boston")
head(Boston)
```

| Variable | Description                            |
|----------|----------------------------------------|
| `medv`   | Median home value (\$1,000s)           |
| `rm`     | Average rooms per dwelling             |
| `lstat`  | \% lower-status population             |
| `crim`   | Per capita crime rate                  |
| `age`    | \% owner-occupied units built pre-1940 |

------------------------------------------------------------------------

## Why These Three Variables?

**Business question:**\
*"How do neighborhood characteristics jointly affect property values?"*

-   `rm` → quality / size signal
-   `lstat` → socioeconomic signal\
-   `medv` → outcome (price)

A real estate firm, bank, or urban planner would care deeply about this relationship.

------------------------------------------------------------------------

## Setting Up in R

```{r}
# Install packages (only needed once)
#install.packages("plotly")
#install.packages("MASS")

# Load libraries
library(plotly)
library(MASS)

# Load dataset
data("Boston")

# Quick check
dim(Boston)      # 506 rows, 14 columns
summary(Boston[, c("rm", "lstat", "medv")])
```

------------------------------------------------------------------------

## 3D Scatter Plot — Code

```{r}
#| echo: false
fig <- plot_ly(
  data   = Boston,
  x      = ~rm,
  y      = ~lstat,
  z      = ~medv,
  type   = "scatter3d",
  mode   = "markers",
  marker = list(
    size       = 4,
    color      = ~medv,
    colorscale = "Viridis",
    showscale  = TRUE
  )
)

fig <- fig %>% layout(
  title = "Boston Housing: Rooms, Status & Home Value",
  scene = list(
    xaxis = list(title = "Avg Rooms (rm)"),
    yaxis = list(title = "Lower Status % (lstat)"),
    zaxis = list(title = "Median Value ($1000s)")
  )
)

fig
```

------------------------------------------------------------------------

## 3D Scatter Plot — What We See

**Key observations:**

-   **High `rm` + Low `lstat`** → High home values (yellow cluster at the top, approaching \$50k)
-   **Low `rm` + High `lstat`** → Low home values (dark purple cluster at the bottom, around \$10k)
-   The **Viridis color scale** reinforces the z-axis — yellow = high value, purple = low value
-   The data forms a **diagonal ridge** rising from high `lstat`/low `rm` toward low `lstat`/high `rm`
-   A small cluster of **very high value homes** (\$45–50k) appears isolated at low `lstat` and high `rm`, suggesting a premium market segment

::: callout-note
Try rotating the chart in the HTML version to see the full structure!
:::

------------------------------------------------------------------------

## 3D Surface Plot — Code

```{r}
# Fit linear model
model <- lm(medv ~ rm + lstat, data = Boston)

# Build prediction grid
rm_seq    <- seq(min(Boston$rm),    max(Boston$rm),    length.out = 50)
lstat_seq <- seq(min(Boston$lstat), max(Boston$lstat), length.out = 50)
grid      <- expand.grid(rm = rm_seq, lstat = lstat_seq)
grid$pred <- predict(model, newdata = grid)
z_matrix  <- matrix(grid$pred, nrow = 50, ncol = 50)

# Surface plot
surface_fig <- plot_ly(
  x = ~rm_seq, y = ~lstat_seq, z = ~z_matrix,
  type = "surface", colorscale = "RdBu"
) %>%
  layout(
    title = "Predicted Home Value Surface",
    scene = list(
      xaxis = list(title = "Avg Rooms"),
      yaxis = list(title = "Lower Status %"),
      zaxis = list(title = "Predicted Value")
    )
  )

surface_fig
```

------------------------------------------------------------------------

## Surface Plot — Business Interpretation

The surface shows the **predicted median home value** across all combinations of `rm` and `lstat`

**How a business uses this:**

| Business            | Use Case                                  |
|---------------------|-------------------------------------------|
| 🏠 Real Estate Firm | Identify undervalued vs premium zones     |
| 🏦 Bank / Lender    | Assess collateral risk by neighborhood    |
| 🏙️ Urban Planner    | Target investment in low-value areas      |
| 📊 Investment Fund  | Screen property acquisition opportunities |

------------------------------------------------------------------------

## Business Applications of 3D Visualization

**Beyond real estate:**

-   **Customer Segmentation** — Age × Income × Spending Score
-   **Financial Risk** — Two risk factors × Portfolio loss
-   **Supply Chain** — Lead time × Order qty × Cost
-   **Marketing** — Campaign spend × Channel × Conversion rate

::: callout-important
The key is always the same: use 3D when **three variables together** tell a story that two cannot.
:::

------------------------------------------------------------------------

## Key Takeaways

✅ 3D visualization reveals **multivariate relationships** hidden in 2D

✅ `plotly` in R makes it **interactive and easy** to implement

✅ The Boston Housing dataset provides a **realistic business context**

✅ Always ask: *does the third dimension add genuine insight?*

✅ Use **interactive HTML** — never static 3D for print

------------------------------------------------------------------------

## References

::: {#refs}
:::

------------------------------------------------------------------------

## Thank You

**Questions?**

Amine Elouafi\
elouafi.amine\@stud.hs-fresenius.de\
Matriculation Number: 401109492 
